---
title: "NPP, GPP, and Rh Processing and Inital Analysis"
output: html_notebook
---

## Input Data 
```{r, message = FALSE, warning=FALSE}
# I wanted to use data.table but there were issues with the grouping by and aggregating. 
library(dplyr) 
library(tidyr)
library(ggplot2)
library(magrittr)
library(data.table)
library(lubridate)
library(tibble)
# install.packages('udunits2')
# cmip6_annual_mean <- drake::readd(cmip6_annual_mean)
# cmip6_global_mean <- drake::readd(cmip6_global_mean)
out_dir <- "CMIP6"
# raw_data <- drake::readd(cmip6_raw)
```

Because the lat and lon coordinates are formatted funky we are going to want to modify them so that they are easier to work with. 
```{r}
raw_data <- read.csv('~/Documents/PNNL/landmodels/CMIP6/yr_latlon_fldmean.csv', stringsAsFactors = FALSE)
# Subset the data so  that it only contains data from the years we are intrested in. 
raw_data <- raw_data[raw_data$year %in% 1980:2015, ]
# It looks like CNRM-CM6-1 reports the units in kg CO2 while the other models reported the values in kg of C. 
# Convert the CNRM-CM6-1 output to be consistent with the other models. 
raw_data <- as.data.table(raw_data)
raw_data <- raw_data[model == "CNRM-CM6-1" & variable %in% c('rh', 'npp'), value := value * (12.0107/44.01)]
raw_data <- raw_data[model == "CNRM-CM6-1-HR" & variable %in% c('rh', 'npp'), value := value * (12.0107/44.01)]
raw_data <- as_tibble(raw_data)

# sum(is.na(raw_data))

if(sum(is.na(raw_data)) > 0){
  raw_data <- raw_data[!is.na(raw_data$value), ]
}

# Create a latitude column, start by parsing out the coordinate information
# from the cdo argument column and format as a matrix. 
coords <- strsplit(gsub(pattern = '-sellonlatbox,', replacement = '', x = raw_data$cdo_arg), split = ',')
coords <- matrix(unlist(coords), nrow = length(coords), byrow = TRUE)[,3]
raw_data[['coords']] <- as.numeric(coords)
raw_data$value_rate <- raw_data$value
raw_data$value <- raw_data$value * raw_data$area # take area into consider
raw_data$units <- "kg s-1" # take area into consider
```

Save a data frame of the area per coordinates. 
```{r}
df_area_coords <- raw_data[ , names(raw_data) %in% c("raw_data", "experiment", "model", "coords", "area", "units")]
df_global_area <- df_area_coords %>%
  group_by(model, experiment) %>%  
  summarise(area = sum(area)) %>%  
  ungroup()
```

Calculate the annual mean flux, in theory the dimensions should not change because this is annual data. This is a precautionary step  just incase the results for a single year were split over mulitple files. 
```{r}
raw_data %>%  
  group_by(year, variable, experiment, ensemble, model, coords, units) %>%  
  summarise(value = mean(value),
            value_rate = mean(value_rate)) %>% 
  ungroup(.) -> 
  annual_mean
```

Since the units are currently the average kg C m^-2 s^-1 for a year let's convert them to g C m^-2 yr^-1

```{r}
# First multiply the rate by the number of seconds in a year.
annual_mean$value <- annual_mean$value * 3.154e7
annual_mean$value_rate <- annual_mean$value_rate * 3.154e7

# Now convert from kg to g. 
annual_mean$value <- udunits2::ud.convert(x = annual_mean$value, u1 = 'kg', u2 = 'g')
annual_mean$value_rate <- udunits2::ud.convert(x = annual_mean$value_rate, u1 = 'kg', u2 = 'g')
annual_mean$units <- gsub(x = annual_mean$units, pattern = "s-1", replacement = "yr-1")
annual_mean$units <- "g yr-1"
```

Calculate the global rate. 

```{r}
annual_mean %>%  
  group_by(year, variable, experiment, ensemble, model, units) %>%  
  summarise(value = sum(value)) %>% 
  ungroup(.) -> 
  global_mean
```

Make some diagnositc plots to make sure that. The results make sense and are consistent. 

```{r}
global_mean %>% 
  ggplot() + 
  geom_line(aes(year, value, color = model, group = interaction(ensemble,  variable, model))) + 
  facet_wrap('variable')
```


```{r}
global_mean %>%  
  mutate(keep  = if_else(variable == 'gpp' & model %in% c("E3SM-1-1", "E3SM-1-1-ECA"), FALSE, TRUE)) %>% 
  mutate(keep = if_else(variable == 'rh' & model %in% c("E3SM-1-1", "E3SM-1-1-ECA"), FALSE, keep)) %>%  
  filter(keep) %>% 
  dplyr::select(-keep) %>% 
  dplyr::filter(value > -1) -> 
  clean_global_data
  
clean_global_data %>% 
  ggplot() + 
  geom_line(aes(year, value, color = model, group = interaction(ensemble,  variable, model))) + 
  facet_wrap('variable') + 
  labs(title = 'Selected Annual Global Data', y = "g yr-1") + 
  theme_bw()
```


```{r}
clean_global_data %>% 
  dplyr::select(year, variable, experiment, ensemble, model, units) %>% 
  distinct() -> 
  data_to_keep
  
clean_laltlon_data <- inner_join(annual_mean, data_to_keep) 
```

Make a quick plot of the lat / lon data for a single year. 

```{r}
clean_laltlon_data %>%  
  filter(year == '2012') %>% 
  ggplot(aes(coords, value_rate, color = model, group = interaction(model, ensemble))) + 
  geom_line() + 
  facet_wrap('variable', scales = 'free') + 
  labs(title = 'C Flux by Lat and Lon', subtitle = '2012', y = "g yr-1") + 
  theme_bw()

clean_laltlon_data %>%  
  filter(year == '2012') %>% 
  ggplot(aes(coords, value, color = model, group = interaction(model, ensemble))) + 
  geom_line() + 
  facet_wrap('variable', scales = 'free') + 
  labs(title = 'C Flux by Lat and Lon', subtitle = '2012', y = "g yr-1") + 
  theme_bw()
```

## Calculate the Percent of Global Total. 

```{r}
global <- rename(clean_global_data, global = value)

percent_data <- left_join(clean_laltlon_data, global, by = c("year", "variable", "experiment", 
                                                            "ensemble", "model", "units")) %>%  
  mutate(percent = 100 * (value / global)) %>%  
  rename(latlon = value)


percent_data %>% 
  filter(year == 2012) %>% 
  ggplot() + 
  geom_line(aes(coords, percent, color = model, group = interaction(ensemble, model))) +
  facet_wrap('variable')

```

## Format 
Now that our data looks decent let's save the data. For simplicity's sake let's save a copy of each of the time series by variable. 
```{r}
yrs <- paste0(range(clean_global_data$year), collapse = '-')
gpp_data <- clean_global_data[clean_global_data$variable == 'gpp', ]
npp_data <- clean_global_data[clean_global_data$variable == 'npp', ]
rh_data  <- clean_global_data[clean_global_data$variable == 'rh', ]

# write.csv(x = gpp_data, file = file.path(out_dir, paste0('gpp_global_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = npp_data, file = file.path(out_dir, paste0('npp_global_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = rh_data, file = file.path(out_dir, paste0('rh_global_', yrs, '.csv')), row.names = FALSE)
```

Next save a copy of the raw lat and lon values. 
```{r}
yrs <- paste0(range(clean_laltlon_data$year), collapse = '-')
gpp_data <- clean_laltlon_data[clean_laltlon_data$variable == 'gpp', ]
npp_data <- clean_laltlon_data[clean_laltlon_data$variable == 'npp', ]
rh_data  <- clean_laltlon_data[clean_laltlon_data$variable == 'rh', ]

# write.csv(x = gpp_data, file = file.path(out_dir, paste0('gpp_latlon_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = npp_data, file = file.path(out_dir, paste0('npp_latlon_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = rh_data, file = file.path(out_dir, paste0('rh_latlon_', yrs, '.csv')), row.names = FALSE)
```


```{r}
yrs <- paste0(range(percent_data$year), collapse = '-')
gpp_data <- percent_data[percent_data$variable == 'gpp', ]
npp_data <- percent_data[percent_data$variable == 'npp', ]
rh_data  <- percent_data[percent_data$variable == 'rh', ]

# write.csv(x = gpp_data, file = file.path('/Users/jian107/Documents/PNNL/landmodels/CMIP6', paste0('gpp_latlon_percent_global_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = npp_data, file = file.path('/Users/jian107/Documents/PNNL/landmodels/CMIP6', paste0('npp_latlon_percent_global_', yrs, '.csv')), row.names = FALSE)
# write.csv(x = rh_data, file = file.path('/Users/jian107/Documents/PNNL/landmodels/CMIP6', paste0('rh_latlon_percent_global_', yrs, '.csv')), row.names = FALSE)
```


Save information about the area per sampled region. 
```{r}
df_area_coords$units <- 'm^2'
# write.csv(x = df_area_coords, file = file.path(out_dir, paste0('area_latlon_model.csv')), row.names = FALSE)
```


## Print Session Info
```{r}
sessionInfo()
```


